Generalized Plackett-Luce Likelihoods

### 2022-09-29 inst/j_sports_econ_F1points.tex
abstract must be <= 100 words according to submission website

\newpage
\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
\section*{Abstract}
The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors, with changes being
controversial.  The inaugural points system was $(8,6,4,3,2)$, that
is, 8 points to the winner, 6 to second place, and so on and in 2020
the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
overall ranking of the competitors is determined by summing points
awarded over a season, but this system suffers from two defects:
firstly, the points are intrinsically arbitrary in the sense that the
points system does not affect the competitors' behaviour but can
change the overall ranking.  Secondly, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
including no arbitrary points system and being amenable to statistical
testing of nulls.  I go on to assess a number of reasonable points
systems objectively, by comparing likelihood-based rankings with
points-based rankings.  In one well-defined statistical sense a Borda
points system is preferred, and in another a ``winner takes all''
system is best.
\newpage

@@

We thus see a comparison between two ordering systems.  Taking 2017 as
an example, drivers Hamilton and Vettel are respectively first and
second according to both ranking ranking procedures; but \raik\ and
Bottas are third and fourth, and fourth and third, according to points
and likelihood respectively. We define the \emph{degree of agreement}
between the two ranking systems as the maximal value of \(r\) such
that places \(1,2,\ldots, r\) all match.  Thus, from
Figure~\ref{orderingbypointsandlikelihood}, the degree of agreement
between likelihood ranking and points ranking for the years 2016-2019
would be 11,1,2,2 respectively.

>>>>

\newpage
\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
\section*{Abstract}
%The points scoring system of Formula 1 motor racing is a long-standing
%contention among fans and competitors, with changes being
%controversial.  The inaugural points system was $(8,6,4,3,2)$, that
%is, 8 points to the winner, 6 to second place, and so on and in 2020
%the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
%overall ranking of the competitors is determined by summing points
%awarded over a season, but this system suffers from two defects:
%firstly, the points are intrinsically arbitrary in the sense that the
%points system does not affect the competitors' behaviour but can
%change the overall ranking.  Secondly, it is difficult to assess
%statistical nulls using accumulated points.  Here, I use Plackett-Luce
%likelihood to identify a ranking for the competitors, this method
%including no arbitrary points system and being amenable to statistical
%testing of nulls.  I go on to assess a number of reasonable points
%systems objectively, by comparing likelihood-based rankings with
%points-based rankings.  In one well-defined statistical sense a Borda
%points system is preferred, and in another a ``winner takes all''
%system is best.

The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors.  The inaugural points system
was $(8,6,4,3,2)$, that is, 8 points to the winner, 6 to second place,
and so on; in 2020 the points system was changed to
$(25,18,15,12,10,8,6,4,2,1)$.  However, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
being amenable to statistical testing.  I go on to assess a
number of reasonable points systems objectively: in one well-defined
statistical sense a Borda points system is optimal.

\newpage

@@

We thus see a comparison between two ordering systems.  Taking 2017 as
an example, drivers Hamilton and Vettel are respectively first and

second according to both ranking ranking procedures; but \raik\ and
Bottas are third and fourth, and fourth and third, according to points
and likelihood respectively. We define the \emph{degree of agreement}
between the two ranking systems as the maximal value of \(r\) such
that places \(1,2,\ldots, r\) all match.  Thus, from
Figure~\ref{orderingbypointsandlikelihood}, the degree of agreement
between likelihood ranking and points ranking for the years 2016-2019
would be 11,1,2,2 respectively.
